---
title: "Literature"
output: rmarkdown::html_vignette
bibliography: literature.bib  
link-citations: true
csl: literature_style.csl
vignette: >
  %\VignetteIndexEntry{Literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  message = FALSE
)
```


```{r, echo = FALSE, include = FALSE}
# Description of how to add new references:
# Simply add them in the literature.bib file in BibTeX style
# For further details see here:
# https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html

# Citation Style
# Is defined in literature_style.csl it is based on American Psychological
# Association 7th edition:
# https://editor.citationstyles.org/styleInfo/?styleId=http%3A%2F%2Fwww.zotero.org%2Fstyles%2Fapa # nolint
# Additionally titles are printed in bold (edited with edit style)
```


## This bibliography contains all peer-reviewed publications linked to WATLAS

The `.bib` file containing the full bibliography in `BibTeX` can be downloaded [here](https://github.com/allertbijleveld/tools4watlas/blob/master/vignettes/literature/literature.bib).

### Introduction 

Tracking animal movement is crucial for understanding interactions with changing environments and predicting the effects of anthropogenic activities, particularly in ecologically significant areas like the Wadden Sea. The WATLAS system  (Wadden Sea Advanced Tracking and Localisation of Animals in real life Systems) enables high-resolution monitoring of small bird movements, offering insights into space use, individual variation, and social networks, thereby supporting research and conservation efforts in the region. A detailed description of WATLAS can be found in **@bijleveld2022**.

### Validation

Validation of methods is crucial to understand the strength and limitations of a method.In **@beardsworth2022** we tested the accuracy and precision of WATLAS using concurrent GPS measurements as a reference.

### Methods

A pipeline with coding examples for cleaning high-throughput tracking data with `atlastools` (`tools4watlas` builds on `atlastools`) is presented in **@gupte2022**.

### Ecology

**@ersoy2022** & **@ersoy2024** 


## Bibliography

---
nocite: '@*'
---

