---
title: "Literature"
output: rmarkdown::html_vignette
bibliography: literature.bib  
link-citations: true
csl: literature_style.csl
vignette: >
  %\VignetteIndexEntry{Literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  message = FALSE
)
```


```{r, echo = FALSE, include = FALSE}
# Description of how to add new references:
# Simply add them in the literature.bib file in BibTeX style
# For further details see here:
# https://bookdown.org/yihui/rmarkdown-cookbook/bibliography.html

# Citation Style
# Is defined in literature_style.csl it is based on American Psychological
# Association 7th edition:
# https://editor.citationstyles.org/styleInfo/?styleId=http%3A%2F%2Fwww.zotero.org%2Fstyles%2Fapa # nolint
# Additionally titles are printed in bold (edited with edit style)
```


## This bibliography contains all peer-reviewed publications linked to WATLAS

The `.bib` file containing the full bibliography in `BibTeX` can be downloaded [here](https://github.com/allertbijleveld/tools4watlas/blob/master/vignettes/literature/literature.bib).

### Introduction 

Tracking animal movement is crucial for understanding interactions with changing environments and predicting the effects of anthropogenic activities, particularly in ecologically significant areas like the Wadden Sea. The WATLAS system  (Wadden Sea Advanced Tracking and Localisation of Animals in real life Systems) enables high-resolution monitoring of small bird movements, offering insights into space use, individual variation, and social networks, thereby supporting research and conservation efforts in the region. A detailed description of WATLAS can be found in **@bijleveld2022**.

In **@nathan2022** we discuss how big-data approaches, such as high-throughput tracking with WATLAS, can lead to an increased understanding of the ecology of animal movement. Particularly that advances in high-throughput wildlife tracking systems now allow more thorough investigation of variation among individuals and species across space and time, the nature of biological interactions, and behavioral responses to the environment.  

### Prologue to WATLAS

ATLAS builds on and is inspired by 'Time Of Arrival'-tracking developed by **@maccurdy2009**. This tracking method is described in **@maccurdy2019** that also includes a pilot study in the Wadden Sea as a 'proof of concept'. After these initial tests in 2008-2009, this pre-WATLAS TOA-system was deployed in the Wadden Sea with 15 receivers studying red knot habitat use. After much trial and error, we could show in **@bijleveld2016** how red knots selected habitat to maximise their energy intake rates. They did, however, not select areas with the highest density of prey but trade-off prey quantity and quality. Moreover, individuals differed in how they leaned towards quantity or quality of prey in selecting mudflat habitat. 

After the success in the Wadden Sea, we also deployed TOA-tracking in Mauritania. Consistent with previous empirical studies, patch residence times in the field were positively correlated with gizzard mass. The slope of this correlation, as well as the observed range of patch residence times, was in accordance with the simulated values. The 12 canutus Red Knots with experimentally reduced gizzard masses did not decrease patch residence times in response to the reduction in gizzard mass. *These findings suggest that diet preferences can indeed cause the observed among-individual variation in gizzard mass and patch residence times [@RN2322]*.

Having deployed the same tracking and resouce sampling methods in Mauritania and the Wadden Sea allowed a comparisson within species [@RN2675]. Compared to Banc d’Arguin, resource patches in the Wadden Sea were larger and the maximum local resource abundance was higher. However, because of constraints set by digestive capacity, the average potential intake rates by red knots were similar at the two study sites. Space-use patterns differed as predicted from these differences in resource landscapes. Whereas foraging red knots in the Wadden Sea roamed the mudflats in high aggregation without site fidelity (i.e. grouping nomads), at Banc d’Arguin they showed less aggregation but were strongly site-faithful (i.e. solitary residents).

As part of a broader review on circadian rhythms, we reveal individual differences in tidal and circadian foraging rhythms of red knots and highlight potential fruitful avenues for further studies [@RN2594].

### Validation

Validation of methods is crucial for understanding the strengths and limitations. In **@beardsworth2022** we tested the accuracy and precision of WATLAS using concurrent GPS measurements as a reference. The median accuracy of WATLAS was 4 m compared with GPS localizations. Localizations that were collected by more receiver stations were more accurate. The three-receiver localizations provided an accuracy of 10 m, which increased to 3 m with seven receivers contributing to the localization. Applying Filter-Smoothing on the data further increased the accuracy to 6 m for three-receiver localizations and to 2 m for seven-receiver localizations. 

### Methods

A pipeline with coding examples for cleaning (e.g. Filter-Smoothing) high-throughput tracking data with `atlastools` (`tools4watlas` builds on `atlastools`) is presented in **@gupte2022**.

In  **@toledo2022** we describe our tags and particulary the design of these versatile, widely-applicable, and field-proven 'Vildehaye' tags for wildlife sensing and radio tracking. Also, we discuss longevity of tags and show that WATLAS tags with a CR2032 battery transmitting at 6 s can last 226 days. 

### Ecology

#### Migration, relocation and departure decisions  
Using WATLAS to record the timing of migration, we showed that red knots that had ad libitum food in captivity departed earlier from the Wadden Sea on spring migration than birds with restricted food access. Becasue Red Knots adjust mass gain and moult rates to local foraging conditions, this study suggests that improved food conditions at staging sites, like the Wadden Sea, could enable earlier departures and help migratory birds better track advancing spring under climate warming [RN3310].

In **@gobbens2024**, we studied the environmental conditions that red knots selected for relocation flights within their wintereing range across the North Sea to the United Kingdom. Approximately 37% of tagged red knots departed yearly and on average did so a few hours after sunset, 4h before high tide, with tailwinds, and little cloud cover.  

#### Habitat use
Red knots foraging on the thick-shelled cockles, that are swallowed whole, needed to trade-off quantity and quality of their prey. In **@bijleveld2016** we show how they, therefore, select habitat with intermediate prey densities to maximise energy intake rates. Other shorebirds, do select for highest desnities where prey are easiliy discovered. In **Penning et al. (in prep)** we show how Sanderling select habitat with the highest desnities of shrimp.

#### Individual variation
In **@ersoy2022** we showed that foraging tactics and diet are associated with the personality trait exploration, independent of morphological differences. WATLAS was used to locate tagged individuals on mudflats for detailed behavioural observations.  

Following this result, in **@ersoy2024** we studied the development of consistent exploration behaviour and found that juvenile red knots had a more diverse diet than adults and had less consistent personalites. We discuss a pathway how early foraging experiences could shape development of exploratory personalities. WATLAS was used to show how juvenile red knots differed in habitat use, which is presented in the appendix. 

## Bibliography

---
nocite: '@*'
---

