---
title: "Plotting data"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  message = FALSE
)
```

This vignette shows different ways on how to plot WATLAS data.

## Using `ggplot2` and a basemap layer

We can create simple base maps of the study area by using the function `atl_create_bm()`. This function uses a data.table with x- and y-coordinates to check the required bounding box (which can be extended with a buffer in meters) and spatial features (polygons) of land, lakes, mudflats and rivers. Without adding costmized data, it will default to the data provided in the package. By changing `asp` the desired aspect ratio can be chosen (default is "16:9"). If no data are provided the function creates a map around Griend with a specified buffer. 

#### Load packages and data

```{r}
library(tools4watlas)
library(data.table)
library(ggplot2)
library(viridis)
library(RColorBrewer)

# Load example data
data <- data_example
```

#### Map with points and tracks

```{r, fig.cap = "Points and tracks on basemap"}

# Create base map
bm = atl_create_bm(data, buffer = 800)

# Plot points and tracks
bm +
  geom_path(data = data, aes(x, y, colour = tag), alpha = 0.1, 
            show.legend = FALSE) +
  geom_point(data = data, aes(x, y, colour = tag), size = 0.5, 
             show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") # for up to 8 categories
  # scale_color_viridis_d(option = "H") # for N categories 

```

#### Heatmap of all positions

```{r, fig.cap = "Heatmap of all positions"}
library(scales)

# Round data to 200 m grid cells
data_heatmap <- copy(data)
data_heatmap[, c("x_round", "y_round") := list(
  plyr::round_any(x, 200),
  plyr::round_any(y, 200)
)]
data_heatmap <- data_heatmap[, .N, by = c("x_round", "y_round")]

# Plot heatmap
bm + 
  geom_tile(data = data_heatmap, aes(x_round, y_round, fill = N), linewidth = 0.1, show.legend = T) +
  scale_fill_viridis(
    option = "A", discrete = FALSE, trans = "log10", name = "N positions",
    breaks = trans_breaks("log10", function(x) 10^x), 
    labels = trans_format("log10", math_format(10^.x)), 
    direction = -1)
```

#### Interactive leaflet map with `mapview`

This is useful when manually checking specific data. Check for more details the [mapview website](https://r-spatial.github.io/mapview/index.html). Nopte that one can change the base map by clicking in the layer symbol, to for example a satellite image. 

```{r, fig.height = 2, fig.cap = "Interactive map with mapview()"}
library(mapview)
library(dplyr)
library(sf)

# make data spatial
d_sf <- atl_as_sf(data, additional_cols = c("tag", "datetime"))

# add track
d_sf_lines <- d_sf %>% 
  group_by(tag) %>%
  summarise(do_union = FALSE) %>%
  st_cast("LINESTRING")

# Plot interactive map
mapview(d_sf_lines, zcol = "tag", legend = FALSE) + mapview(d_sf, zcol = "tag")
```


#### Static map with `OpenStreetMap` 



```{r, fig.cap = "Static map with satellite image"}
library(OpenStreetMap)
library(sf)

# make data spatial and transform projection to WGS 84 (used in osm)
d_sf <- atl_as_sf(data, additional_cols = c("tag", "datetime"))
d_sf <- st_transform(d_sf, crs = st_crs(4326)) 

# get bounding box
bbox <- atl_bbox(d_sf, asp = "16:9", buffer = 500)

# extract openstreetmap 
# other 'type' options are "osm", "maptoolkit-topo", "bing", "stamen-toner",
# "stamen-watercolor", "esri", "esri-topo", "nps", "apple-iphoto", "skobbler";
osm_map <- openmap(c(bbox["ymax"], bbox["xmin"]),
                   c(bbox["ymin"], bbox["xmax"]),
                   type = "bing", mergeTiles = TRUE) 

bm <- autoplot.OpenStreetMap(osm_map)

# transform points to Mercator and add transformed coordinates to data
d_sf <- st_transform(d_sf, crs = st_crs(3857)) 
osm_coords <- st_coordinates(d_sf)
data[, `:=`(x_osm = osm_coords[, 1], y_osm = osm_coords[, 2])]

# plot
bm +
  geom_path(data = data, aes(x_osm, y_osm, colour = tag), alpha = 0.1, 
            show.legend = FALSE) +
  geom_point(data = data, aes(x_osm, y_osm, colour = tag), size = 0.5, 
             show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Longitude", y = "Latitude") +
 coord_sf(crs = 3857)
```


