---
title: "Smooth and thin data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Smooth and thin data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png"
)
```

 This vignette shows how to smooth and thin WATLAS data. 

```{r, message = FALSE}
# Packages
library(tools4watlas)
library(ggplot2)

# Path to csv with filtered data
data_path <- system.file(
  "extdata", "watlas_data_filtered.csv",
  package = "tools4watlas"
)

# Load data
data <- fread(data_path, yaml = TRUE)

```


## Median smooth data

To reduce error in the position data, a basic smoother such as a median filter can be applied. the function `atl_median_smooth` calculates the median coordinates within a window of positions set by `moving window`.  

```{r echo=T, results='hide'}		
# Smooth the data
data <- atl_median_smooth(data, moving_window = 5)
```

The resulting table overwrites the smoothed coordinates in the columns `x` and `y` and keeps the original ones in the columns `x_raw` and `y_raw`.

### Calculate speed and turning angle

After median filtering the data, the speeds need to be recalculated. We will also calculate turning angles. 

Note: the distance between median smoothed positions can be 0 and therefore will produce NAs and a warning

```{r echo=T}		
# Recalculate speed
data <- atl_get_speed(data, type = c("in", "out"))
```

#### Look at the data

This plot just shows one example of a raw and median smooted track.

```{r, fig.cap = "Smoothed track (black) on top of raw track (red)", fig.align = "center"}	
# subset first tag
data_subset <- data[tag == data[1]$tag]

# subset some data to look at
from <- min(data_subset[, datetime]) + 1 * 3600
to <- min(data_subset[, datetime]) + 12 * 3600
data_subset <- data_subset[datetime %between% c(from, to)]

# Create basemap
bm <- atl_create_bm(data_subset)

# Plot
bm +
  geom_path(
    data = data_subset, aes(x_raw, y_raw),
    color = "firebrick3", linewidth = 0.5
  ) +
  geom_path(
    data = data_subset, aes(x, y),
    color = "black", linewidth = 0.5
  ) +
  geom_point(
    data = data_subset, aes(x_raw, y_raw),
    color = "firebrick3", size = 1.2
  ) +
  geom_point(
    data = data_subset, aes(x, y),
    color = "black", size = 1
  )
```
## Save data for the next steps

```{r, eval = FALSE}
# Save data
fwrite(
  data,
  file = "../inst/extdata/watlas_data_smoothed.csv", yaml = TRUE
)
```

## Thin data

Depending on the desired analysis, it might make sense to thin data, either by aggregation or by subsampling. Both methods return fixed time steps (depending on the interval). 

### By aggregation

Returns the mean of all columns for each time step. The additional column `n_aggregated` shows how many positions were aggregated for this position. Time and datetime are returned rounded down to the desired interval.

```{r echo=T}		
# Thin the data by aggregation with a 60-second interval
thinned_aggregated <- atl_thin_data(
  data = data,
  interval = 60,
  id_columns = c("tag", "species"),
  method = "aggregate"
)

# Show head of selected data
head(thinned_aggregated[, .(tag, time, datetime, x, y, n_aggregated)]) |>
  knitr::kable(digits = 2)
```

### By subsampling

Returns the first position for each time step. The column `n_subsampled` shows from how many positions this position was sampled.

```{r echo=T}		
# Thin the data by subsampling with a 60-second interval
thinned_subsampled <- atl_thin_data(
  data = data,
  interval = 60,
  id_columns = c("tag", "species"),
  method = "subsample"
)

# Show head of selected data
head(thinned_subsampled[, .(tag, time, datetime, x, y, n_subsampled)]) |>
  knitr::kable(digits = 2)
```
