---
title: "Add residency patches"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add residency patches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

# Work in progress

This vignette shows how to assign residency patches with WATLAS data. 

Ignores different sampling interals and species specific residency patch parameters at the moment.

### Load packages and required data

```{r, message = FALSE}
# packages
library(tools4watlas)
library(ggplot2)
library(viridis)
library(foreach)
library(doFuture)

# load example data
data <- data_example

# file path to WATLAS teams data folder
fp <- atl_file_path("watlas_teams")

# load tide pattern data
tide_data <- fread(paste0(
  fp, "waterdata/allYears-tidalPattern-west_terschelling-UTC.csv"
))
```

## Calculate residency patches by tag

Subset relevant columns only (otherwise slows down the process).

```{r echo=T}		
# unique tag ID
id <- unique(data$tag)

# loop by tag ID
patch_data <- foreach(i = id, .combine = "rbind") %do% {
  atl_res_patch(
    data[tag == i, .(tag, posID, time, datetime, x, y, speed_in)],
    max_speed = 3, lim_spat_indep = 75, lim_time_indep = 180,
    min_fixes = 3, min_duration = 120
  )
}

# extract data from all residency patches
patch_data_ext <- patch_data[, cbind(
  rbindlist(lapply(patchdata, function(x) cbind(x))), duration
), by = .(tag, patch)]

# merge back with data
data <- merge(
  data, patch_data_ext[, .(tag, posID, patch, duration)],
  by = c("tag", "posID"), all.x = TRUE
)
```

## Plot by tag and tide ID

### Example for one tag and tide

```{r, fig.width = 8.89, fig.height = 5.2}
atl_check_res_patch(
  tag = "3038", tide = "2023513",
  data, patch_data, tide_data
)
```

### Loop through all tags and tides


```{r, eval = FALSE}
# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# unique ID combinations
idc <- unique(data[, c("species", "tag", "tideID")])

# loop to make plots for all
foreach(i = seq_len(nrow(tag_tide))) %dofuture% {

  # plot and save for each combination
  atl_check_res_patch(
    tag = idc$tag[i], tide = idc$tideID[i],
    data, patch_data, tide_data,
    filename = paste0(
      "./outputs/res_patch_check/",
      idc$species[i], "_tag_", idc$tag[i], "_tide_", idc$tideID[i]
    )
  )

}

# close parallel workers
plan(sequential)
```

## Plot residency patches (Example red knot)

```{r}
# duration in minutes
patch_data[, duration := duration / 60]

# subset red knot
data_subset <- data[tag == 3038]
patch_data_subset <- patch_data[tag == 3038]

# create basemap
bm <- atl_create_bm(data_subset, buffer = 500)

# track with residency patches coloured
bm +
  geom_path(data = data_subset, aes(x, y), alpha = 0.1) +
  geom_point(
    data = data_subset, aes(x, y), color = "grey",
    show.legend = FALSE
  ) +
  geom_point(
    data = data_subset[!is.na(patch)], aes(x, y, color = as.character(patch)),
    size = 1.5, show.legend = FALSE
  )

# plot residency patches itself by duration
bm +
  geom_point(
    data = patch_data_subset,
    aes(x_median, y_median, color = duration, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_viridis()
```

## Plot residency patches (with species colours)

```{r}
# create basemap
bm <- atl_create_bm(data, buffer = 500)

# add species
du <- unique(data, by = "tag")
patch_data <- patch_data[du, on = "tag", `:=`(species = i.species)]

# plot residency patches itself by duration and species
bm +
  geom_point(
    data = patch_data,
    aes(x_median, y_median, color = species, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_manual(
    values = atl_spec_cols(),
    labels = atl_spec_labs("multiline"),
    name = ""
  )
```
