---
title: "Add residence patches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add residence patches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

In this vignette, we provide a general workflow to group WATLAS position data into so-called 'residence patches'. Note that the parameter settings can/should be adapted for different species' behaviour and data quality.  

## Background and parameter explanation
The `atl_res_patch()` function is designed to segment and aggregate WATLAS movement data into residence patches. The main parameter is speed (`max_speed`). With perfect data that would be the only parameter necessary to adjust, because the speed flying, walking or standing do not overlap. Because WATLAS data have localization error (comparable to GPS, see [Beardsworth et al. 2022](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.13913)) and gaps when birds are out of range of receivers, we need to have additional variables for classifying these data into robust residence patches.

The logic of the function is to first identify proto-patches (preliminary residence patches). When subsequent positions have a speed smaller than `max_speed`, a distance smaller than `lim_spat_indep` and a time gap smaller than `lim_time_indep`, they are assigned to the same proto patch. Proto-patches with fewer than `min_fixes` positions are filtered out. 

For each proto patch, the median position is calculated as well as the time between two subsequent proto-patches (time between last location and first location of the next proto-patch). Proto patches are merged into residence patches, if the distance between the median positions of two subsequent proto-patches is smaller than `lim_spat_indep` and the time between the proto-patch is less then `lim_time_indep`. 

Lastly, a unique patch ID is assigned to each residence patch ordered by time from 1 to n.

Note that without cleaning the data or having short intervals (e.g. 3 sec), position error can lead to speed outliers, which affects the creation of proto-patches. When calculating residence patches, it is therefor recommended to first filter (`var_max < 5000`) and smooth (`moving_window = 5`) the data. Depending on the species' behaviour or quality of the data, the `max_speed` can also be set to a higher value, or the data can be thinned first. Keep in mind that smoothing and thinning will influence the speeds between positions, and thus the creation of residence patches.

**Parameter overview:**

   - **`max_speed`:** A numeric value specifying the maximum speed (m/s) between two subsequent positions that would be considered non-transitory.
   - **`lim_spat_indep`:** A numeric value specifying the maximum distance (m) between subsequent residence patches for them to be considered independent. In combination with `lim_time_indep`, this parameter avoids making a new proto-patch from gaps in the data when the bird was actually still at the same location.
   - **`lim_time_indep`:** A numeric value specifying the time difference (min) between two subsequent residence patches for them to be considered independent. In combination with `lim_spat_indep`, this parameter can prevent the creation new proto-patches when there are large gaps in the data. For example, at the roost site a bird might not move for a long time at a location with poor coverage by receivers. If the bird then moves away and sends data from the same position, we can assume that all missed positions were also at this place.
   - **`min_fixes`:** The minimum number of positions for proto-patches. To make sure that residence patches have at least a few positions.
   - **`min_duration`:** The minimum duration (s) for classifying residence patches. With a high-sampling interval (e.g. 1 s), short residence patches can be created, which are not biological relevant.

**Guidelines to choose parameters:**

In general, when deciding on the optimal parameters, the key is to find a good balance between true and false positives. We give some starting points for each parameter, which has to be adjusted based on the data quality and species behaviour.

   - **`max_speed`:** Should be as high as possible in between walking and flying speeds. A good starting point is **3 m/s**, but could be reduced if it prevents the creation of proto-patches, or increased if too many proto-patches are created. Having too many proto-patches is not alwasy an issue becasue subsequent proto-patches can be merged if the distance between them is small enough (set by `lim_spat_indep`). Likewise, errors in positiong data can inflate the creation of proto patches, but these will be corrected in the procedure.
   - **`lim_spat_indep`:** Provides the distance between two proto-patches at which they are considered independent. This is a key parameter, because it prevents the creation of new proto-patches when the bird is still at the same location. A good starting point is **75 m**, but could be increased if the position data has many/large gaps, or if the species has elongated foraging patches.
   - **`lim_time_indep`:** Typically, **180 min** (3 hours) works fine, but could be increased with position data that has large gaps. For example, when the analysis is focused on roosting behaviour, this could even be increased to e.g. 6 hours (360 min), to deal with large gaps in the data that can occur with roosting birds not moving and being at the same place with bad signal strength.
   - **`min_fixes`:** This should be set as small as possible and **3 positions** usually works. Setting this variable >1 allows assigning proto patches only if subsequent positions are consistently above the 1 `max_speed` and `lim_spat_indep` and not just once becasue of an outlier, for instance. 
   - **`min_duration`:** Should be set as small as possible while maintaining most biological relevance. A value of **120 sec** (2 minutes) seems reasonable.

## Load packages and required data

```{r, message = FALSE}
# packages
library(tools4watlas)
library(ggplot2)
library(viridis)
library(foreach)
library(doFuture)

# load example data
data <- data_example

# file path to WATLAS teams data folder
fp <- atl_file_path("watlas_teams")

# load tide pattern data
tidal_pattern <- fread(paste0(
  fp, "waterdata/allYears-tidalPattern-west_terschelling-UTC.csv"
))
```

## Calculate residence patches by tag

To reduce the memory size for parallel computing, we will first subset the relevant columns from the data. This can be skipped for small data tables. We will then run `atl_res_patch()` for each tag ID in parallel. The column `patch` is added to the data table, which provides the assigned patch ID's for the positions.  

```{r echo=T, warning=FALSE}		
# subset relevant columns
data <- data[, .(species, posID, tag, time, datetime, x, y, tideID)]

# extract the unique tag IDs
id <- unique(data$tag)

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop through all tags to calculate residence patches
data <- foreach(i = id, .combine = "rbind") %dofuture% {
  atl_res_patch(
    data[tag == i],
    max_speed = 3, lim_spat_indep = 75, lim_time_indep = 180,
    min_fixes = 3, min_duration = 120
  )
}

# close parallel processing
plan(sequential)

# show head of the summary table
head(data) |> knitr::kable(digits = 2)
```

## Evaluate residence patch classification and parameters

The function `atl_check_res_patch()` can be used to evaluate the residence patch classification by tag and tide ID. the function plots the track with residence patches on a map and shows the duration (time in a patch in min) as coloured polygon on the map and as plot against the time in a separate plot. Time starts on the top and goes from high tide to high tide (solid blue lines), as well as indicating low tide (dashed blue line). The title of the plot gives basic information about the data and the water level for the corresponding tide. 

### Inspect one tag and tide

We can select one tag and tide to plot. Additionally, we need to specify the offset for the tidal data we use (e.g. 30 min for West-Terschelling) and a buffer (in m) around the residence patch data to create the polygon. This buffer should be set to half of `lim_spat_indep` (maximum distance between subsequent residence patches at which they will be considered independent), ensuring that the polygons around residence patches correspond to the spatial distance threshold used to merge residence patches. 

```{r, fig.align = "center", fig.alt = "Overview plot res patches one tide", fig.width = 10, fig.height = 6}
atl_check_res_patch(
  data[tag == "3038"], tide_data = tidal_pattern,
  tide = "2023513", offset = 30,
  buffer_res_patches = 75 / 2
)
```

### Inpect many tags and tides

To get a general overview, we can also loop through and plot all data by tag and tide, or for example a random sample of 100 tags and tides. The plots are saved in any directory (e.g. `./outputs/res_patch_check/`), which has to be created before running the code.

```{r, eval = FALSE}
# create table with data combinations to plot
idc <- unique(data[, c("species", "tag", "tideID")])

# sample 100 combinations to plot
set.seed(123)
idc <- idc[sample(.N, 100)]

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop to make plots for all
foreach(i = seq_len(nrow(idc))) %dofuture% {

  # plot and save for each combination
  atl_check_res_patch(
    data[tag == idc$tag[i]],
    tide_data = tidal_pattern,
    tide = idc$tideID[i], offset = 30,
    buffer_res_patches = 75,
    filename = paste0(
      "./outputs/res_patch_check/",
      idc$species[i], "_tag_", idc$tag[i], "_tide_", idc$tideID[i]
    )
  )

}

# close parallel processing
plan(sequential)
```

Based on these plots and perhaps additional checks, the parameters can be adjusted to improve the classification of residence patches.

## Summary of residence patch data

Once satisfied with the residence patch classification, we can summarize the residence patches by tag and patch ID and merge the desired columns back to our full data table. 

```{r}
# summary of residence patches
data_summary <- atl_res_patch_summary(data)

# standardise duration to minutes
data_summary[, duration := duration / 60]

# merge desired summary columns with original data table
data[data_summary, on = c("tag", "patch"), `:=`(
  duration = i.duration,
  disp_in_patch = i.disp_in_patch
)]

# show head of the summary table
head(data_summary) |> knitr::kable(digits = 2)
```

| Column            | Description                                                                 |
|------------------|------------------------------------------------------------------------------|
|**tag**           | 4 digit tag ID (character), i.e. last 4 digits of the full tag number        |
|**patch**         | Patch ID                                                                     |
|**nfixes**        | Number of fixes in the patch                                                 |
|**x_mean**        | Mean X-coordinate in meters (UTM 31 N)                                       |
|**x_median**      | Median X-coordinate in meters (UTM 31 N)                                     |
|**x_start**       | X-coordinate at the start of the residence patch (UTM 31 N)                  |
|**x_end**         | X-coordinate at the end of the residence patch (UTM 31 N)                    |
|**y_mean**        | Mean Y-coordinate in meters (UTM 31 N)                                       |
|**y_median**      | Median Y-coordinate in meters (UTM 31 N)                                     |
|**y_start**       | Y-coordinate at the start of the residence patch (UTM 31 N)                  |
|**y_end**         | Y-coordinate at the end of the residence patch (UTM 31 N)        			  |
|**time_mean**     | Mean UNIX time (seconds) of the positions in the residence patch             |
|**time_median**   | Median UNIX time (seconds) of the positions in the residence patch           |
|**time_start**    | Start UNIX time (seconds) of the patch                                       |
|**time_end**      | End UNIX time (seconds) of the patch                                         |
|**dist_in_patch** | Total distance (in meters) travelled within the patch (cummulative distance) |
|**dist_bw_patch** | Distance (in meters) between end of previous and start of current patch      |
|**time_bw_patch** | Time (in seconds) between end of previous and start of current patch         |
|**disp_in_patch** | Straight-line displacement (in meters) between start and end of the patch    |
|**duration**      | Time duration (in seconds) between first and last position in patch          |



## Plot the residence pathes
It might also be useful to plot the residence patches using `ggplot2`. 

### Plot by tag
In this example, we will plot the residence patches for one red knot (tag 3038). In the frist example, the residence patches are coloured by patch ID. To show the full track, the transient (unassigned) positions are plotted in grey.

```{r, fig.alt = "residence patches within track colored by ID"}
# subset red knot
data_subset <- data[tag == 3038]
data_summary_subset <- data_summary[tag == 3038]

# create basemap
bm <- atl_create_bm(data_subset, buffer = 500)

# track with residence patches coloured
bm +
  geom_path(data = data_subset, aes(x, y), alpha = 0.1) +
  geom_point(
    data = data_subset, aes(x, y), color = "grey",
    show.legend = FALSE
  ) +
  geom_point(
    data = data_subset[!is.na(patch)], aes(x, y, color = as.character(patch)),
    size = 1.5, show.legend = FALSE
  )
```

In the second example, the residence patches are plotted at their median positions with the size and colour scaled to their duration (in minutes).

```{r, fig.alt = "residence patches by duration in patch"}
# plot residence patches itself by duration
bm +
  geom_point(
    data = data_summary_subset,
    aes(x_median, y_median, color = duration, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_viridis()
```

### Plot by species

Similarly, we can plot the residence patches by species. For this, we need to merge the species information back to the summary table for residence patches. The residence patches are coloured by species and scaled by duration (in minutes).

```{r, fig.alt = "residence patches colored by species"}
# create basemap
bm <- atl_create_bm(data, buffer = 500)

# add species
du <- unique(data, by = "tag")
data_summary <- data_summary[du, on = "tag", `:=`(species = i.species)]

# plot residence patches itself by duration and species
bm +
  geom_point(
    data = data_summary,
    aes(x_median, y_median, color = species, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_manual(
    values = atl_spec_cols(),
    labels = atl_spec_labs("multiline"),
    name = ""
  )
```
