---
title: "Add residency patches"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add residency patches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

# Work in progress

This vignette shows how to assign residency patches with WATLAS data. 

```{r, message = FALSE}
# Packages
library(tools4watlas)
library(ggplot2)
library(viridis)
library(foreach)

# Path to csv with filtered data
data_path <- system.file(
  "extdata", "watlas_data_filtered.csv",
  package = "tools4watlas"
)

# Load data
data <- fread(data_path, yaml = TRUE)

```


## Median smooth data

To further reduce error in the localization data, a basic smoother such as a median filter can be applied. The resulting table has the smoothed location stored in the `x` and `y` column and the original location stored in the `x_raw` and `y_raw` column. 

```{r echo=T, results='hide'}		
# Smooth the data
data <- atl_median_smooth(data, moving_window = 5)
```

### Recalculate speed

After smoothing the data, the speeds need to be recalculated. We now also calculate turning angles. 
Note: the distance between smoothed positions can be 0 and therefore will produce NAs and a warning

```{r echo=T}		
# Recalculate speed
data <- atl_get_speed(data, type = c("in", "out"))
```

## Calculate residency patches by tag

```{r echo=T}		
# unique tag ID
id <- unique(data$tag)

# loop by tag ID
rp <- foreach(i = id, .combine = "rbind") %do% {
  atl_res_patch_speed(
    data[tag == i],
    max_speed = 3, lim_spat_indep = 50, lim_time_indep = 180,
    min_fixes = 3, min_duration = 180
  )
}

# extract data from all residency patches
drp <- rp[, rbindlist(lapply(
  patchdata, function(x) cbind(x)
)), by = .(tag, patch)]

# merge back with data
data <- merge(
  data, drp[, .(tag, posID, patch)],
  by = c("tag", "posID"), all.x = TRUE
)
```

## Plot residency patches (Example red knot)

```{r}
# duration in minutes
rp[, duration := duration / 60]

# subset red knot
data_subset <- data[tag == 3038]
rp_subset <- rp[tag == 3038]

# create basemap
bm <- atl_create_bm(data_subset, buffer = 500)

# track with residency patches coloured
bm +
  geom_path(data = data_subset, aes(x, y), alpha = 0.1) +
  geom_point(
    data = data_subset, aes(x, y), color = "grey",
    show.legend = FALSE
  ) +
  geom_point(
    data = data_subset[!is.na(patch)], aes(x, y, color = as.character(patch)),
    size = 1.5, show.legend = FALSE
  )

# plot x against datetime
ggplot() +
  geom_line(
    data = data_subset, aes(datetime, x), color = "grey",
    show.legend = FALSE
  ) +
  geom_point(
    data = data_subset[!is.na(patch)],
    aes(datetime, x, color = as.character(patch)),
    size = 1.5, show.legend = FALSE
  ) +
  theme_bw()

# plot residency patches itself by duration
bm +
  geom_point(
    data = rp_subset,
    aes(x_median, y_median, color = duration, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_viridis()
```

## Plot residency patches (with species colours)

```{r}
# create basemap
bm <- atl_create_bm(data, buffer = 500)

# add species
du <- unique(data, by = "tag")
rp <- rp[du, on = "tag", `:=`(species = i.species)]

# plot residency patches itself by duration and species
bm +
  geom_point(
    data = rp,
    aes(x_median, y_median, color = species, size = duration),
    show.legend = TRUE, alpha = 0.5
  ) +
  scale_color_manual(
    values = atl_spec_cols(),
    labels = atl_spec_labs("multiline"),
    name = ""
  )
```
