---
title: "Interpolate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png"
)
```

This article shows how to interpolate between positions thus filling gaps in the WATLAS data. Under some circumstances this can be useful, but one should be cautious and mindful of the data and research question. 

For some analyses, gaps in positioning data can bias the results and interpolating positions across small gaps can be necesarry. For example, if birds move some distance within residence patches, the median position in the patch does not properly reflect space use. Moreover, within residence patches, especially when birds are on the ground roosting, it is likely that birds do not move much and interpolation between positions can be done safely. 

Here, we show how to interpolate and fill small data gaps within residence patches. 

## Load packages and data

```{r, message = FALSE}
# packages
library(tools4watlas)
library(ggplot2)
library(viridis)
library(foreach)
library(doFuture)

# load example data
data <- data_example
```

## Prepare data

To interpolate positions within residence patches, we first need to calculate residence patches as described in the article ["Add residence patches"](https://allertbijleveld.github.io/tools4watlas/articles/extended_workflow/add_residence_patches.html). 

For convenience and to avoid interpolating many (superfluous) positions, we suggest aggregating the data into regular intervals as described in the article ["Smooth and thin data"](https://allertbijleveld.github.io/tools4watlas/articles/smooth_and_thin_data.html). 

```{r echo=T, warning=FALSE}		
# subset relevant columns
data <- data[, .(species, posID, tag, time, datetime, x, y, tideID)]

# extract the unique tag IDs
id <- unique(data$tag)

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop through all tags to calculate residence patches
data <- foreach(i = id, .combine = "rbind") %dofuture% {
  atl_res_patch(
    data[tag == i],
    max_speed = 3, lim_spat_indep = 75, lim_time_indep = 180,
    min_fixes = 3, min_duration = 120
  )
}

# close parallel processing
plan(sequential)

# summary of residence patches
data_summary <- atl_res_patch_summary(data)

# thin the data to 1 min interval
data <- atl_thin_data(
  data = data,
  interval = 60,
  id_columns = c("tag", "species"),
  method = "aggregate"
)
```

### Check data

Before interpolation, it is good inspect the residence patch summary and check whether interpolating positions is appropriate. For example, if the residence patches are overly elongated there might be a mistake in the residence patch calculation. Likewise, if there are large spatial or temporal gaps between positions, interpolating might be inappropriate.

```{r echo=T, warning=FALSE}
# merge species to data_summary
data_summary <- merge(data_summary, unique(data[, .(tag, species)]), by = "tag")

# look at the maximal time spent within residence patches and
# the maximal distance between the start and end position of residence patches
data_summary[, .(
  max_duration = max(duration) |> atl_format_time(),
  max_dist_start_end = max(dist_start_end)
), by = species]
```

The oystercatcher spent 7.6 h in a residence patch, which is quite long, but possible.

## Interpolate between positions

In this example, we trust the residence patch calculation and we want to interpolate the data to 1 min intervals within residence patches. We set the `max_gap` to 8 h, which is longer than the longest residence patch duration, so that all residence patches are interpolated and we set `patches_only` to `TRUE`, so that only data within residence patches are interpolated.

```{r echo=T, warning=FALSE}
# interpolate data to 1 min intervals - only within residence patches
data_int <- atl_interpolate_track(
  data = data,
  tag = "tag",
  x = "x",
  y = "y",
  time = "time",
  patch = "patch",
  interp_interval = 60,
  max_gap = 3600 * 8,
  max_dist = NULL,
  patches_only = TRUE
)

# show head of the table
head(data_int) |> knitr::kable(digits = 2)
```

### Plot data and check interpolation

Here, we plot one tag as an example. The positions are coloured by residence patch, and the interpolated points are shown in black.

```{r echo=T, warning=FALSE}
# subset one tag
data_subset <- data_int[tag == "3288"]

# create basemap
bm <- atl_create_bm(data_subset, buffer = 800)

# plot points and tracks with standard ggplot colours
bm +
  geom_path(
    data = data_subset, aes(x, y),
    linewidth = 0.5, alpha = 0.1, show.legend = FALSE
  ) +
  geom_point(
    data = data_subset, aes(x, y, colour = patch),
    size = 1, alpha = 1, show.legend = FALSE
  ) +
  geom_point(
    data = data_subset[interpolated == TRUE], aes(x, y), color = "black",
    size = 0.5, alpha = 1, show.legend = FALSE
  )
```

Looks good!
