---
title: "Animate data"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Animate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
``` 

This article gives a simple workflow on how to animate movement data. It works with any type of basemap and allows to adjust everything as wanted. It first shows how to animate data with a simple basemap and then how to do it with a bathymetry data basemap and the water level. The first two steps and the last are the same (load packages, prepare movement data, make animation).

Install `mapmate` from GitHub: `remotes::install_github("leonawicz/mapmate")`.

#### Load packages

```{r}
# packages
library(tools4watlas)
library(ggplot2)
library(viridis)
library(foreach)
library(doFuture)
library(ragg)
library(mapmate)

# additional to add water data
library(terra)
library(tidyterra)
library(sf)
```

## Prepare movement data

Set the folder path where the png's are created, load the data, create time steps with a desired interval (e.g. 10 min), delete existing files.

```{r}
# file path
path <- "C:/Users/jkrietsch/temp/animation"

# load example data
data <- data_example

# create time steps
ts <- atl_time_steps(
  datetime_vector = data$datetime,
  time_interval = "10 min",
  output_path = path
)

# delete existing files (if any)
unlink(paste0(path, "/*"), recursive = TRUE)
```

# Simple animation of movements

## Check plot

Make a basemap as desired and plot with all data to check the outcome, scalebar and time stamp. Adjust everything as desired (check with saving png in the defined size).

```{r, fig.alt = "Check the basemap", fig.align = "center"}
# create basemap
bm <- atl_create_bm(data, buffer = 800)

# plot points and tracks to check
bm +
  geom_path(
    data = data, aes(x, y, colour = tag),
    linewidth = 0.5, alpha = 0.1, show.legend = FALSE
  ) +
  geom_point(
    data = data, aes(x, y, colour = tag),
    size = 0.5, alpha = 1, show.legend = FALSE
  ) +
  # add time stamp
  annotate(
    "text",
    x = -Inf, y = -Inf, hjust = -0.1, vjust = -2.4,
    label = paste0(format(data[1]$datetime, "%Y-%m-%d %H:%M")), size = 4
  )
```

## Loop to create png's for each step

Create a png for each step (check steps - that is how many png's are created). Run first with an example step to check (e.g. `step <- 50`) outcome for one png. Depending on the number of png's, run first on a subset to check if everything is as desired, once everything is fine, run for all steps. A desired color scale can be simply added to `p`.

In the subset step, the maximal tail length can be chosen (here, 6 h). Then we use `atl_alpha_along()` to create an fading alpha and `atl_size_along()` to make a comet shape - adjust parameters as desired (see function description for more details). Afterwards we add this path to the basemap and add the time stamp. 

```{r, eval = FALSE}
# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# steps
steps <- seq_len(nrow(ts))

# loop to create pngs for each time step
foreach(i = steps) %dofuture% {

  # define time step
  time_step <- ts[i]$datetime # current date

  # subset data
  ds <- data[datetime %between% c(time_step - 3600 * 6, time_step)]

  # create alpha and size
  if (nrow(ds) > 0) {
    ds[, a := atl_alpha_along(
      datetime,
      head = 30, skew = -2
    ), by = tag]
  }
  if (nrow(ds) > 0) {
    ds[, s := atl_size_along(
      datetime,
      head = 70, to = c(0.3, 2.5)
    ), by = tag]
  }

  # add tracks to basemap
  p <- bm +
    geom_path(
      data = ds, aes(x, y, color = tag), alpha = ds$a, linewidth = ds$s,
      lineend = "round", show.legend = FALSE
    ) +

    # add time stamp
    annotate(
      "text",
      x = -Inf, y = -Inf, hjust = -0.1, vjust = -2.4,
      label = paste0(format(time_step, "%Y-%m-%d %H:%M")), size = 4
    )

  # save png
  agg_png(
    filename = ts[i, path],
    width = 3840, height = 2160, units = "px", res = 300
  )
  print(p)
  dev.off()

}
```

# Animation of movements with water level

## Add tide and bathymetry data

We add water level data to `ts` to know the waterlevel at each step and crop the bathymetry data to the extend of the map (use same buffer as for basemap).  

```{r}
# file path to WATLAS teams data folder
fp <- atl_file_path("watlas_teams")

# sub path to tide data
tidal_pattern_fp <- paste0(
  fp, "waterdata/allYears-tidalPattern-west_terschelling-UTC.csv"
)
measured_water_height_fp <- paste0(
  fp, "waterdata/allYears-gemeten_waterhoogte-west_terschelling-clean-UTC.csv"
)

# load tide data
tidal_pattern <- fread(tidal_pattern_fp)
measured_water_height <- fread(measured_water_height_fp)

# add unix time
ts[, time := as.numeric(datetime)]

# add tide data to movement data
ts <- atl_add_tidal_data(
  data = ts,
  tide_data = tidal_pattern,
  tide_data_highres = measured_water_height,
  waterdata_resolution = "10 minute",
  offset = 30
)

# file path to Birds, fish 'n chips GIS/rasters folder
fp <- atl_file_path("rasters")

# load bathymetry data
bat <- rast(paste0(fp, "bathymetry/2024/bodemhoogte_20mtr_UTM31_int.tif"))

# bbox (should be buffer used for basemap)
bbox <- atl_bbox(data, buffer = 800)

# crop the raster using the bounding box
bat_c <- crop(bat, bbox)

# wrap to use in parallel loop
bat_w <- wrap(bat_c)
```

## Check plot

Make a basemap as desired and plot with all data to check the outcome, scalebar and time stamp. Adjust everything as desired (check with saving png in the defined size).

```{r, fig.alt = "Check the basemap with bathymetry data", fig.align = "center"}
# create basemap
bm <- atl_create_bm(
  data,
  buffer = 800, raster_data = bat_c, option = "batymetry", scalebar = FALSE
)

# extract example water level
threshold <- 0 / 100 # water level at the time (/100 to scale to m)
bat_m <- bat_c < threshold # mask below threshold (TRUE = 1)
bat_m[bat_m == 0] <- NA # remove land
waterline <- as.polygons(bat_m, values = TRUE, dissolve = TRUE) |>
  st_as_sf() # extract polygon with water level

# check plot with all data
bm +
  # add water level
  geom_sf(
    data = waterline, fill = "lightblue", alpha = 0.4,
    color = scales::alpha("white", 0.1), linewidth = 3
  ) +
  # add points and tracks
  geom_path(
    data = data, aes(x, y, colour = tag),
    linewidth = 0.5, alpha = 0.1, show.legend = FALSE
  ) +
  geom_point(
    data = data, aes(x, y, colour = tag),
    size = 0.5, alpha = 1, show.legend = FALSE
  ) +
  # add time stamp
  annotate(
    "text",
    x = -Inf, y = -Inf, hjust = -0.1, vjust = -2.4,
    label = paste0(format(data[1]$datetime, "%Y-%m-%d %H:%M")), size = 4
  ) +
  # add scale bar
  ggspatial::annotation_scale(
    aes(location = "br"),
    text_cex = 1, height = unit(0.3, "cm"),
    pad_x = unit(0.4, "cm"), pad_y = unit(0.6, "cm")
  ) +
  # set extend again (overwritten by geom_sf)
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]), expand = FALSE
  )
```

## Loop to create png's for each step

Same as above just with added water level polygon and scale bar (needs to be added above water).

```{r, eval = FALSE}
# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# steps
steps <- seq_len(nrow(ts))

# loop to create pngs for each time step
foreach(i = steps) %dofuture% {

  # define time step
  time_step <- ts[i]$datetime # current date

  # subset data
  ds <- data[datetime %between% c(time_step - 3600 * 6, time_step)]

  # create alpha and size
  if (nrow(ds) > 0) {
    ds[, a := atl_alpha_along(
      datetime,
      head = 30, skew = -2
    ), by = tag]
  }
  if (nrow(ds) > 0) {
    ds[, s := atl_size_along(
      datetime,
      head = 70, to = c(0.3, 2.5)
    ), by = tag]
  }

  # extract water level
  bat_c <- unwrap(bat_w)
  threshold <- ts[i]$waterlevel / 100 # water level at the time (scale to m)
  bat_m <- bat_c < threshold # mask below threshold (TRUE = 1)
  bat_m[bat_m == 0] <- NA # remove land
  waterline <- as.polygons(bat_m, values = TRUE, dissolve = TRUE) |>
    st_as_sf() # extract polygon with water level

  # add everything to the basemap
  p <- bm +
    # add water level
    geom_sf(
      data = waterline, fill = "lightblue", alpha = 0.4,
      color = scales::alpha("white", 0.1), linewidth = 3
    ) +
    # add track
    geom_path(
      data = ds, aes(x, y, color = tag), alpha = ds$a, linewidth = ds$s,
      lineend = "round", show.legend = FALSE
    ) +
    # add time stamp
    annotate(
      "text",
      x = -Inf, y = -Inf, hjust = -0.1, vjust = -2.4,
      label = paste0(format(time_step, "%Y-%m-%d %H:%M")), size = 4
    ) +
    # add scale bar
    ggspatial::annotation_scale(
      aes(location = "br"),
      text_cex = 1, height = unit(0.3, "cm"),
      pad_x = unit(0.4, "cm"), pad_y = unit(0.6, "cm")
    ) +
    # set extend again (overwritten by geom_sf)
    coord_sf(
      xlim = c(bbox["xmin"], bbox["xmax"]),
      ylim = c(bbox["ymin"], bbox["ymax"]), expand = FALSE
    )

  # save png
  agg_png(
    filename = ts[i, path],
    width = 3840, height = 2160, units = "px", res = 300
  )
  print(p)
  dev.off()

}
```

## Make a animation using `ffmpeg` via `mapmate`

Adjust the frame rate (`rate`) as desired (depending on the time step interval). File is created in the same path, but this can also be changed as desired.

```{r, eval = FALSE}
# make animation
ffmpeg(
  dir = path, output_dir = path, pattern = atl_ffmpeg_pattern(ts[1]$path),
  output = "Animation.mp4", rate = 8, details = TRUE, overwrite = TRUE
)
```
