---
title: "Create a basemap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create a basemap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

This article shows how to make different basemaps for plotting movement data. We present three options for creating a basemap using: 1) polygons delineating areas of mudflat, water and land provided within the package (see also the Article [Basemap data](https://allertbijleveld.github.io/tools4watlas/articles/package_development/basemap_data.html)), 2) bathymetry data, 3) backgrounds from the package `maptiles`, and 4) the package `OpenStreetMap`.

## Load packages

```{r}
# packages
library(tools4watlas)
library(ggplot2)
```

## Using polygons

One can create a simple basemaps of the study area by using the function `atl_create_bm()`. This function uses a data.table with x- and y-coordinates to check the required bounding box (which can be extended with a buffer in meters) and spatial features (polygons) of land, lakes, mudflats and rivers. Without adding spatial data, it will default to the spatial data provided in the package. By changing `asp` the desired aspect ratio can be chosen (default is "16:9"). The resulting map is always in EPSG:32631 (WGS 84 / UTM zone 31N), but the data can be provided in other projections, which then needs to be specified as `projection`.

### Basemap with extent of movement data
Instead of using a priori coordinates to construct a basemap, one can also use the movement data to set the extent of the map. 

```{r, fig.alt = "Points and tracks on basemap"}
# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 1000)

# plot
bm +
  geom_point(data = data, aes(x, y), alpha = 0.1, color = "darkorange")
```

### Basemap independent of movement data

This can be useful when one wants to zoom into a specific area of the plot or has an area of interest, but the movement data can extent beyond this specifieds range. If no data are provided, the function defaults to a map around Griend (our main study site) with a specified buffer. 

```{r, fig.alt = "Basemap around Griend"}
# create basemap
bm <- atl_create_bm(buffer = 30000)

# plot
bm
```

Alternatively, one can provide a table with one or multiple locations, which will then be used to make the map. This can for example be a location in EPSG:4326 (WGS 84 in degrees) that can be extracted from [Google Maps](https://www.google.com/maps/@53.1892078,5.5670405,68619m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI1MDIwNS4xIKXMDSoASAFQAw%3D%3D) by a right click on a specific location. Here I choose a point a bit east of Griend, to include Richel and Ballastplaat in the map.

```{r, fig.alt = "Basemap around Griend"}
# define location
location <- data.table(x = c(5.275), y = c(53.2523))

# create basemap
bm <- atl_create_bm(location, buffer = 7000, projection = sf::st_crs(4326))

# plot
bm
```

Instead of using one central point and a buffer we can also provide two points (corners of a rectangle) to specify the basemap. When we do this, we also might not want to use a buffer and a specific aspect ratio, so we can set `buffer = 0` and `asp = NULL`, but if desired we can also use a buffer and an aspect ratio. The example below shows how to create a basemap around Griend and Richel with two corner points, which is useful when we want to zoom into a specific area of the map. The first point is north-west of Richel and the second point south-east of Griend.

```{r, fig.alt = "Basemap around Griend"}
# specify corner points
corner_points <- data.table(x = c(5.107, 5.330), y = c(53.303, 53.230))

# create basemap
bm <- atl_create_bm(
  corner_points,
  buffer = 0, asp = NULL, projection = sf::st_crs(4326)
)

# plot
bm
```

`bm` is a standard `ggplot2` object and can be manipulated as one. For example, we can overwrite the default theme to add axis labels and ticks.

```{r, fig.alt = "Basemap with axis labels"}
# create basemap
bm <- atl_create_bm(buffer = 10000)

# add axis labels and ticks
bm +
  labs(x = "Longitude", y = "Latitude") +
  theme(
    axis.title   = element_text(),
    axis.text.x  = element_text(),
    axis.text.y  = element_text(),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line()
  )
```

When adding `sf` object to the basemap, the coordinate system will be overwritten and we need to set the bounding box again. For points or lines, it is best to use `geom_point()` and `geom_path()`, which is also faster than plotting the same data with `geom_sf()` (just make sure the data are in the correct coordinate system -  EPSG:32631 (WGS 84 / UTM zone 31N)). Plotting polygons only works with `geom_sf()` and afterwards the bounding box needs to be set again with `coord_sf()` and the buffer in `atl_create_bm()` is ignored.

Show are different ways on how to set the bounding box.

```{r, fig.alt = "Basemap with polygons"}
# additional packages
library(sf)

# load example data
data <- data_example

# bounding box based on data
bbox <- atl_bbox(data, buffer = 1000)

# bounding box based on specified coordinates in EPSG:4326
bbox <- data.table(x = c(5.107, 5.330), y = c(53.303, 53.230)) |>
  st_as_sf(coords = c("x", "y"), crs = 4326) |>
  st_transform(crs = 32631) |>
  atl_bbox(buffer = 1000)

# bounding box based on polygon
bbox <- atl_bbox(grienderwaard, buffer = 1000)

# create basemap with bounding box
bm <- atl_create_bm(bbox, buffer = 0)

# plot bm with bounding box
bm +
  geom_sf(data = grienderwaard, fill = "transparent", color = "firebrick") +
  # set extend again (overwritten by geom_sf)
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]), expand = FALSE
  )
```

## Using bathymetry data

Here, we use a raster file with bathymetry to construct a basemap. Bathymetry data can be found in the "Birds, fish 'n chips" SharePoint folder: `Documents/data/GIS/rasters/`. To run the script set the file path (`fp`) to the local copy of the folder on your computer. The data can also be downloaded from the [Waddenregister](https://datahuiswadden.openearth.nl/geonetwork/srv/eng/catalog.search#/metadata/JkrvJXasRSGAWU1mJLHUzg).

```{r, fig.alt = "Bathymetry map", warning=FALSE}
# additional packages
library(terra)

# file path to Birds, fish 'n chips GIS/rasters folder
fp <- atl_file_path("rasters")

# load bathymetry data
bat <- rast(paste0(fp, "bathymetry/2024/bodemhoogte_20mtr_UTM31_int.tif"))

# create base map with bathymetry data
bm <- atl_create_bm(
  buffer = 5000, raster_data = bat, option = "bathymetry"
)

# plot
bm
```

We can also add some shading (`shade = TRUE`) to the bathymetry data to highlight the geomorphological structures better. Note that calculating shading can take a while, especially for large maps. We remommend using this option only for polished maps.

```{r, fig.alt = "Bathymetry map with shade", warning=FALSE}
# additional packages
library(terra)

# file path to Birds, fish 'n chips GIS/rasters folder
fp <- atl_file_path("rasters")

# load bathymetry data
bat <- rast(paste0(fp, "bathymetry/2024/bodemhoogte_20mtr_UTM31_int.tif"))

# create base map with bathymetry data
bm <- atl_create_bm(
  buffer = 5000, raster_data = bat, option = "bathymetry", shade = TRUE
)

# plot
bm
```

## Using `maptiles`

The package `maptiles` provides an easy way to create basemaps from different tile providers (e.g., OpenStreetMap, Esri satellite images, Stamen toner, etc.). The function `atl_create_bm_tiles()` wraps around the `maptiles` functions to create a basemap based on movement data or specified bounding boxes and otherwise works like `atl_create_bm()`. It is best to use the standard EPSG:32631 (WGS 84 / UTM zone 31N) to create a bounding box (since this means the data don't need to be transformed internally), but other projections also work. The output maps are, however, always in EPSG:4326, so movement data need to be transformed to be added to the plot. `atl_transform_dt()` can be used to add transformed positions to the data table.

### Use satellite map

```{r, fig.alt = "Static map with satellite image using maptiles", eval = TRUE}

# satellite map and buffer around Griend
bm <- atl_create_bm_tiles(
  buffer = 15000, option = "Esri.WorldImagery", zoom = 12
)

# plot
bm

# example with bbox and adding movement data
data <- data_example

# add transformed coordinates in projection of the base map (EPSG:4326)
data <- atl_transform_dt(data)

# plot points and tracks using transformed coordinates.
bm +
  geom_path(
    data = data, aes(x_4326, y_4326, colour = tag),
    linewidth = 0.5, alpha = 0.1, show.legend = FALSE
  ) +
  geom_point(
    data = data, aes(x_4326, y_4326, colour = tag),
    size = 0.5, alpha = 1, show.legend = FALSE
  ) +
  scale_color_discrete(name = paste("N = ", length(unique(data$tag)))) +
  theme(legend.position = "top")
```

## Using `OpenStreetMap`

A range of basemap options are provided by `OpenStreetMap`. First, we have to transform the data to WGS 84 to extract the basemap with the bounding box, and then transform our data to a Mercator projection to plot the data on top of the map.

Unfortunately, the `type = "bing"` (satellite image) is unstable and does not always work.

```{r, fig.alt = "Static map with satellite image", fig.width = 5, fig.height = 4.72}
# additional packages
library(OpenStreetMap)
library(sf)

# load example data
data <- data_example

# make data spatial and transform projection to WGS 84 (used in osm)
d_sf <- atl_as_sf(data, additional_cols = c("tag", "datetime"))
d_sf <- st_transform(d_sf, crs = st_crs(4326))

# get bounding box
bbox <- atl_bbox(d_sf, asp = "16:9", buffer = 10000)

# extract openstreetmap
# other 'type' options are "osm", "maptoolkit-topo", "bing", "stamen-toner",
# "stamen-watercolor", "esri", "esri-topo", "nps", "apple-iphoto", "skobbler";
map <- openmap(
  c(bbox["ymax"], bbox["xmin"]),
  c(bbox["ymin"], bbox["xmax"]),
  type = "esri", mergeTiles = TRUE
)

bm <- autoplot.OpenStreetMap(map)

# transform points to Mercator and add transformed coordinates to data
d_sf <- st_transform(d_sf, crs = sf::st_crs(3857))
osm_coords <- st_coordinates(d_sf)
data[, `:=`(x_osm = osm_coords[, 1], y_osm = osm_coords[, 2])]

# plot
bm +
  geom_point(
    data = data, aes(x_osm, y_osm), alpha = 0.1, color = "darkorange"
  ) +
  coord_sf(crs = 3857, expand = FALSE) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Adding features to a map 

Here, we show how to add additional features, like the WATLAS logo or receivers, to a map.

### Add WATLAS logo

```{r, fig.alt = "basemap with logo"}
# additional packages
library(ggimage)

# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 1000)

# path to WATLAS logo
logo_path <- system.file(
  "extdata", "watlas_logo.png",
  package = "tools4watlas"
)

# define position based on data (here upper left corner)
# adjust as desired
x_pos <- min(data$x) - 2500
y_pos <- max(data$y)

# create table with image of the logo
di <- data.table(x = x_pos, y = y_pos, image = logo_path)

# plot basemap with logo
bm +
  geom_image(data = di, aes(x, y, image = image), size = 0.2)
```

### Add WATLAS receivers

Receiver data are managed by Allert. They are located in the "WATLAS" SharePoint folder: `Documents/data/`. Either specify the path to your local copy of this folder or add the path for your user in the `atl_file_path()` function.

```{r, fig.alt = "basemap with receivers"}
# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 20000)

# file path to WATLAS teams data folder
fp <- atl_file_path("watlas_teams")

# load receivers data
dr <- readxl::read_excel(
  paste0(fp, "receivers/receiver specifications.xlsx"),
  sheet = "receivers"
) |>
  data.table()

# end date for active receivers as system date
dr[active == "ja", date_removed := as.POSIXct(Sys.time(), tz = "UTC")]

# subset all active in period of tracking data
start <- min(data$datetime)
end <- max(data$datetime)

# Check if each row includes period of tracking data
dr[, includes_data := (date_placed <= end & date_removed >= start)]
dr <- dr[includes_data == TRUE]

# plot basemap with receivers active during tracking period
bm +
  geom_point(data = dr, aes(x, y), pch = 17)
```

One can also add the name of the receiver.

```{r, fig.alt = "basemap with receivers and label"}
# additional packages
library(ggrepel)

# plot basemap with receivers and label
bm +
  geom_point(data = dr, aes(x, y), pch = 17, size = 1.5) +
  geom_text_repel(
    data = dr, aes(x, y, label = location),
    max.overlaps = Inf,
    size = 3,
    point.padding = 0.5,
    segment.color = "grey50"
  )
```
