---
title: "Create a basemap"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create a basemap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

This article shows how to make different basemaps to plot the movement data on. Here we present three options: A simple basemap using polygons provided with the package (see also article: [Basemap data](https://allertbijleveld.github.io/tools4watlas/articles/package_development/basemap_data.html)), a simple basemap based on bathymetry data and flexible approach using the package `OpenStreetMap` (e.g. to plot on a satellite image).

#### Load packages

```{r}
# Packages
library(tools4watlas)
library(ggplot2)
```

## Using `ggplot2` and a basemap layer

One can create a simple basemaps of the study area by using the function `atl_create_bm()`. This function uses a data.table with x- and y-coordinates to check the required bounding box (which can be extended with a buffer in meters) and spatial features (polygons) of land, lakes, mudflats and rivers. Without adding spatial data, it will default to the spatial data provided in the package. By changing `asp` the desired aspect ratio can be chosen (default is "16:9"). The resulting map is always in EPSG:32631 (WGS 84 / UTM zone 31N), but the data can be provided in other projections, which then needs to be specified as `projection`.

### Basemap independent of movement data

This can be useful when one wants to zoom into a specific area of the plot or has an area of interest, but the movement data also go out of this range. If no data are provided the function creates a map around Griend (our main study site) with a specified buffer. 


```{r, fig.alt = "Basemap around Griend"}
# create basemap
bm <- atl_create_bm(buffer = 30000)

# plot
bm
```

Alternatively, one can provide a table with one or multiple locations, which will then be used to buffer the map. This can for example be a location in EPSG:4326 (WGS 84 in degrees) that can be exacted from [Google Maps](https://www.google.com/maps/@53.1892078,5.5670405,68619m/data=!3m1!1e3?entry=ttu&g_ep=EgoyMDI1MDIwNS4xIKXMDSoASAFQAw%3D%3D) by a right click on a specific location. Here I choose a point a bit east of Griend, to include Richel and Ballastplaat in the map.

```{r, fig.alt = "Basemap around Griend"}
# define location
location <- data.table(x = c(5.275), y = c(53.2523))

# create basemap
bm <- atl_create_bm(location, buffer = 7000, projection = sf::st_crs(4326))

# plot
bm
```

### Basemap with extend of movement data

```{r, fig.alt = "Points and tracks on basemap"}
# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 1000)

# plot
bm +
  geom_point(data = data, aes(x, y), alpha = 0.1, color = "darkorange")
```

## Basemap with bathymetry data

Bathymetry data can be found in the "Birds, fish 'n chips" SharePoint folder: `Documents/data/GIS/rasters/`. To run the script set the file path (`wd`) to the local copy of the folder on your computer. The data can also be downloaded from the [Waddenregister](https://datahuiswadden.openearth.nl/geonetwork/srv/eng/catalog.search#/metadata/JkrvJXasRSGAWU1mJLHUzg).

```{r, fig.alt = "Static map with satellite image", warning=FALSE}
# additional packages
library(terra)

# Load bathymetry data
wd <- "C:/Users/jkrietsch/OneDrive - NIOZ/Documents/MAP_DATA/"
bat <- rast(paste0(wd, "bathymetry/2024/bodemhoogte_20mtr_UTM31_int.tif"))

# create base map with bathymetry data
bm <- atl_create_bm(
  buffer = 8000, raster_data = bat, option = "batymetry"
)

# plot
bm
```

## Basemap with `OpenStreetMap`

Provides a range of different basemap options. For this we first have to transform the data to WGS 84 to extract the basemap with the bounding box and then transform our data to a Mercator projection to plot the data on top of the map.

Unfortunately, sometimes the `type = "bing"` (satellite image) does not work.

```{r, fig.alt = "Static map with satellite image", fig.width = 5, fig.height = 4.72}
# additional packages
library(OpenStreetMap)
library(sf)

# load example data
data <- data_example

# make data spatial and transform projection to WGS 84 (used in osm)
d_sf <- atl_as_sf(data, additional_cols = c("tag", "datetime"))
d_sf <- st_transform(d_sf, crs = st_crs(4326))

# get bounding box
bbox <- atl_bbox(d_sf, asp = "16:9", buffer = 10000)

# extract openstreetmap
# other 'type' options are "osm", "maptoolkit-topo", "bing", "stamen-toner",
# "stamen-watercolor", "esri", "esri-topo", "nps", "apple-iphoto", "skobbler";
map <- openmap(
  c(bbox["ymax"], bbox["xmin"]),
  c(bbox["ymin"], bbox["xmax"]),
  type = "osm", mergeTiles = TRUE
)

bm <- autoplot.OpenStreetMap(map)

# transform points to Mercator and add transformed coordinates to data
d_sf <- st_transform(d_sf, crs = sf::st_crs(3857))
osm_coords <- st_coordinates(d_sf)
data[, `:=`(x_osm = osm_coords[, 1], y_osm = osm_coords[, 2])]

# plot
bm +
  geom_point(
    data = data, aes(x_osm, y_osm), alpha = 0.1, color = "darkorange"
  ) +
  coord_sf(crs = 3857, expand = FALSE) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Additional features

How to add a WATLAS logo or receivers to a base map.

### Add WATLAS logo to a basemap

```{r, fig.alt = "basemap with logo"}
# additional packages
library(ggimage)

# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 1000)

# path to WATLAS logo
logo_path <- system.file(
  "extdata", "watlas_logo.png",
  package = "tools4watlas"
)

# define position based on data (here upper left corner)
# adjust as desired
x_pos <- min(data$x) - 2500
y_pos <- max(data$y)

# create table with image of the logo
di <- data.table(x = x_pos, y = y_pos, image = logo_path)

# plot basemap with logo
bm +
  geom_image(data = di, aes(x, y, image = image), size = 0.2)
```

### Add WATLAS receivers to a basemap

Receiver data are managed by Allert. They are located in the "WATLAS" SharePoint folder: `Documents/data/`. Either specify the path to your local copy of this folder or add the path for your user in the `atl_file_path()` function.

```{r, fig.alt = "basemap with receivers"}
# load example data
data <- data_example

# create basemap
bm <- atl_create_bm(data, buffer = 20000)

# file path to WATLAS teams data folder
fp <- atl_file_path("watlas_teams")

# load receivers data
dr <- readxl::read_excel(
  paste0(fp, "receivers/receiver specifications.xlsx"),
  sheet = "receivers"
) |>
  data.table()

# end date for active receivers as system date
dr[active == "ja", date_removed := as.POSIXct(Sys.time(), tz = "UTC")]

# subset all active in period of tracking data
start <- min(data$datetime)
end <- max(data$datetime)

# Check if each row includes period of tracking data
dr[, includes_data := (date_placed <= end & date_removed >= start)]
dr <- dr[includes_data == TRUE]

# plot basemap with receivers active during tracking period
bm +
  geom_point(data = dr, aes(x, y), pch = 17)
```

One can also add the name of the receiver.

```{r, fig.alt = "basemap with receivers and label"}
# additional packages
library(ggrepel)

# plot basemap with receivers and label
bm +
  geom_point(data = dr, aes(x, y), pch = 17, size = 1.5) +
  geom_text_repel(
    data = dr, aes(x, y, label = location),
    max.overlaps = Inf,
    size = 3,
    point.padding = 0.5,
    segment.color = "grey50"
  )
```
