---
title: "Plot data interactively"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot data interactively}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

This article shows different ways on how to plot WATLAS data interactively using `mapview`. This is useful when manually checking specific data. Check for more details the [mapview website](https://r-spatial.github.io/mapview/index.html). Note that one can change the base map by clicking in the layer symbol, to for example a satellite image. Each chunk of code only requires this chunk with loading the data to be run before and is otherwise independent. 

#### Load packages

```{r}
# packages
library(tools4watlas)
library(mapview)
```

## Interactive plot for one individual

Subset the individual of choice and transform it into a `sf` with the additional columns of your choice (can be seen when clicking on the point). Colour the track by selecting the desired parameter as `zcol`.

```{r, fig.height = 2, fig.cap = "Interactive map with mapview()"}
# load example data from one tide
data <- data_example[tideID == "2023513"]

# subset data
data_subset <- data[tag == "3063"]

# make data spatial
d_sf <- atl_as_sf(
  data_subset,
  additional_cols = c("species", "datetime", "speed_in", "nbs", "waterlevel")
)

# add track
d_sf_lines <- atl_as_sf(
  data_subset,
  additional_cols = c("species", "datetime", "speed_in", "nbs", "waterlevel"),
  option = "lines"
)

# plot interactive map
mapview(d_sf_lines, zcol = "speed_in", legend = FALSE) +
  mapview(d_sf, zcol = "speed_in")
```

## Interactive plot for multiple individuals

If one wants to plot a lot of data, it is recommended to thin the data first.

```{r, fig.height = 2, fig.cap = "Interactive map with mapview()"}
# load example data
data <- data_example

# thin the data by subsampling with a 60-second interval
data <- atl_thin_data(
  data = data,
  interval = 60,
  id_columns = c("tag", "species"),
  method = "subsample"
)

# make data spatial
d_sf <- atl_as_sf(data, additional_cols = c("datetime", "species"))

# add track
d_sf_lines <- atl_as_sf(
  data,
  additional_cols = c("species"),
  option = "lines"
)

# plot interactive map
mapview(d_sf_lines, zcol = "tag", legend = FALSE) +
  mapview(d_sf, zcol = "tag")
```

## Interactive plot for multiple species

```{r, fig.height = 2, fig.cap = "Interactive map with mapview()"}
# load example data from one tide
data <- data_example[tideID == "2023513"]

# thin the data by subsampling with a 60-second interval
data <- atl_thin_data(
  data = data,
  interval = 60,
  id_columns = c("tag", "species"),
  method = "subsample"
)

# make data spatial
d_sf <- atl_as_sf(data, additional_cols = c("datetime", "species"))

# add track
d_sf_lines <- atl_as_sf(
  data,
  additional_cols = c("species"),
  option = "lines"
)

# plot interactive map
mapview(d_sf_lines, zcol = "species", legend = FALSE) +
  mapview(d_sf, zcol = "species")
```
