---
title: "Plot data in loop"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot data in loop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

This article shows how to make and save plots in a loop and parallel loop using the package [`foreach`]() and [`doFuture`](). This is useful for example when checking the data by tag ID using `atl_check_tag()`.

#### Load packages

```{r}
# packages
library(tools4watlas)
library(ggplot2)
library(foreach)
library(doFuture)
```

## Plot and save data in loop by tag ID

Example of a simple loop by tag ID and using `atl_check_tag()`.

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag/tag_"

# unique ID (here by tag)
id <- unique(data$tag)

# loop to make plots for all
foreach(i = id) %do% {

  # subset data
  data_subset <- data[tag == i]

  # plot and save data
  atl_check_tag(
    data_subset,
    option = "datetime",
    highlight_first = TRUE, highlight_last = TRUE,
    filename = paste0(path, data_subset[1]$species, "_tag_", i)
  )

}
```

Example to make any type of `ggplot` and save the data with `ragg`.

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag/tag_"

# unique ID (here by tag)
id <- unique(data$tag)

# loop to make plots for all
foreach(i = id) %do% {

  # subset data
  data_subset <- data[tag == i]

  # plot data
  p <- ggplot()

  # save
  agg_png(
    filename = paste0(path, i, ".png"),
    width = 3840, height = 2160, units = "px", res = 300
  )
  print(p)
  dev.off()

}
```


## Plot and save data in parallel loop by tag ID

We can use the same structure, simply replacing `%do%` with `%dofuture%`, which is the key advantage of using `foreach.` The only additional step is setting up parallel processing with `registerDoFuture()` and `plan(multisession)`.

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag/"

# unique ID (here by tag)
id <- unique(data$tag)

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop to make plots for all
foreach(i = id) %dofuture% {

  # subset data
  data_subset <- data[tag == i]

  # plot and save data
  atl_check_tag(
    data_subset,
    option = "datetime",
    highlight_first = TRUE, highlight_last = TRUE
    filename = paste0(path, data_subset[1]$species, "_tag_", i)
  )

}

# close parallel workers
plan(sequential)
```
## Plot and save data in parallel loop by tag ID and tide ID

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag_and_tide/"

# unique ID combinations
idc <- unique(data[, c("species", "tag", "tideID")])

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop to make plots for all
foreach(i = seq_len(nrow(idc))) %dofuture% {

  # subset data
  data_subset <- data[tag == idc$tag[i] & tideID == idc$tideID[i]]

  # plot and save data
  atl_check_tag(
    data_subset,
    option = "datetime",
    highlight_first = TRUE, highlight_last = TRUE,
    filename = paste0(
      path, idc$species[i], "_tag_", idc$tag[i], "_tide_", idc$tideID[i]
    )
  )

}

# close parallel workers
plan(sequential)
```

