---
title: "Plot data in loop"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot data in loop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE
)
```

This article shows how to make and save plots in a loop and parallel loop using the package [`foreach`]() and [`doFuture`](). This is useful for example when checking the data by tag ID using `atl_check_tag()`.

#### Load packages

```{r}
# packages
library(tools4watlas)
library(ggplot2)
library(foreach)
library(doFuture)
library(ragg)
```

## Plot and save data in loop

Example of a simple loop by tag ID and using `atl_check_tag()`.

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag/name_tag_"

# unique ID (here by tag)
id <- unique(data$tag)

# loop to make plots for all
foreach(i = id) %do% {

  # subset data
  data_subset <- data[tag == i]

  # plot data
  p <- atl_check_tag(
    data_subset,
    option = "datetime",
    highlight_first = TRUE, highlight_last = TRUE
  )

  # save
  agg_png(
    filename = paste0(path, i, ".png"),
    width = 3840, height = 2160, units = "px", res = 300
  )
  print(p)
  dev.off()

}
```

## Plot and save data in parallel loop

We can use the exact same structure and just replace `%do%` with `%dofuture%`, which is the benefit for using `foreach` and just have to prepare the parallel processing using `registerDoFuture()` and `plan(multisession)`. 

```{r, eval = FALSE}
# load example data
data <- data_example

# file path
path <- "./outputs/maps_by_tag/name_tag_"

# unique ID (here by tag)
id <- unique(data$tag)

# register cores and backend for parallel processing
registerDoFuture()
plan(multisession)

# loop to make plots for all
foreach(i = id) %dofuture% {

  # subset data
  data_subset <- data[tag == i]

  # plot data
  p <- atl_check_tag(
    data_subset,
    option = "datetime",
    highlight_first = TRUE, highlight_last = TRUE
  )

  # save
  agg_png(
    filename = paste0(path, i, ".png"),
    width = 3840, height = 2160, units = "px", res = 300
  )
  print(p)
  dev.off()

}
```
