---
title: "Basemap data"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basemap data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  fig.width = 8.89, fig.height = 5,
  dpi = 300,
  dev = "ragg_png",
  message = FALSE,
  warning = FALSE
)
```

This script shows how to create the basemap data of `tools4watlas`, which are a land polygon of the Dutch Wadden Sea, the mudflats of North Holland and Friesland and waterbodies on Griend. These data were choosen provide a simple map with relevant data allowing fast plotting. Customized basemap data could be created in a similar way and could additional contain buildings, roads, lakes, rivers etc. All data can be found in the "Birds, fish 'n chips" SharePoint folder: `Documents/data/GIS/shapefiles/`. To run the script set the file path (`fp`) to the local copy of the folder on your computer. 

The `OpenStreetMap` land polygon can also be downloaded from [`osmdata`](https://osmdata.openstreetmap.de/data/land-polygons.html) and the regional data of the Netherlands (here used are North Holland and Friesland) can be downloaded from [`Geofabrik`](https://download.geofabrik.de/europe/netherlands.html). 

#### Load packages and specify path to local data
 
```{r}
# Packages
library(data.table)
library(tools4watlas)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)

# file path to Birds, fish 'n chips GIS/shapefiles folder
fp <- atl_file_path("shapefiles")
```

## Define a bounding box of the Dutch Wadden Sea

First define a bounding box which is used to crop the land polygon data.

```{r, fig.cap="Bounding box around the Dutch Wadden Sea"}
# get data from the Netherlands
netherlands <- ne_countries(
  country = "netherlands", scale = "large", returnclass = "sf"
) |>
  st_transform(crs = st_crs(32631))

# point of Griend (and a bit east)
griend <- st_sfc(st_point(c(5.2525 + 0.6, 53.2523)), crs = st_crs(4326)) |>
  st_transform(crs = st_crs(32631))

# bounding box around Griend
bbox <- atl_bbox(griend, asp = "16:9", buffer = 50000)
bbox_sf <- bbox |> st_as_sfc()

# plot
ggplot() +
  geom_sf(data = netherlands) +
  geom_sf(data = bbox_sf, color = "firebrick3", fill = NA) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"])
  )
```

## Extract the land polygon data from this bounding box

```{r, fig.cap="Cropped land polygon around the Dutch Wadden Sea"}
# load osm land polygon
land_ <- st_read(quiet = TRUE, paste0(
  fp, "open_street_map/land-polygons-complete-4326/land_polygons.shp"
)) |>
  st_transform(crs = st_crs(32631))

# crop data
land <- st_intersection(land_, bbox_sf)

# extract only geometry
land <- land["geometry"]

# union to compress
land <- st_union(land)

# plot
ggplot() +
  geom_sf(data = land) +
  geom_sf(data = bbox_sf, color = "firebrick", fill = NA) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"])
  )
```

### Save data

Includes data in the package, if `tools4watlas` is opened as project.

```{r, eval=FALSE}
# save data
save(land, file = "../../data/land.rda", compress = "xz")
```

## Define a polygon of the Dutch Wadden Sea

To simplify the basemap we only want the mudflats from within the Wadden Sea, otherwise *fclass* also includes other wetlands.  

```{r, fig.cap="Polygon of the Dutch Wadden Sea"}
# load polygon of Wadden sea
wadden_sea <- st_read(quiet = TRUE, paste0(
  fp, "wadden_area_legally/pkb_gebied_derde_nota_waddenzee.shp"
)) |>
  st_transform(crs = st_crs(32631))

# crop with bbox
wadden_sea <- st_intersection(wadden_sea, bbox_sf)

# buffer
ws_buffer <- wadden_sea |> st_buffer(1000)
ws_buffer <- ws_buffer[, c("geometry")]

# check data
ggplot() +
  geom_sf(data = land) +
  geom_sf(data = ws_buffer, color = "firebrick", fill = NA)
```

## Extract mudflats and lakes from within the Wadden Sea

We only take the lakes from Griend to not blow up the data. 

```{r, fig.cap="Final basemap data of the Dutch Wadden Sea"}
# Friesland
lakes_fr <- st_read(quiet = TRUE, paste0(
  fp, "open_street_map/friesland-latest-free.shp/gis_osm_water_a_free_1.shp"
))

# North Holland
lakes_nh <- st_read(quiet = TRUE, paste0(
  fp, "open_street_map/noord-holland-latest-free.shp/gis_osm_water_a_free_1.shp"
))

# North Groningen
lakes_g <- st_read(quiet = TRUE, paste0(
  fp, "open_street_map/groningen-latest-free.shp/gis_osm_water_a_free_1.shp"
))


# merge both and change projection
lakes_ <- rbind(lakes_fr, lakes_nh, lakes_g) |>
  unique(by = "osm_id") |>
  st_transform(crs = st_crs(32631))

# crop data
lakes <- st_intersection(lakes_, ws_buffer)

# subset mudflats
mudflats <- lakes[lakes$fclass == "wetland", ]

# union to compress
mudflats <- st_union(mudflats)

# subset lakes
lakes <- lakes[lakes$fclass == "water", ]

# crop to include just Griend
griend <- st_sfc(st_point(c(5.2525, 53.2523)), crs = st_crs(4326)) |>
  st_transform(crs = st_crs(32631))
bbox_sf <- atl_bbox(griend, asp = "16:9", buffer = 3000) |> st_as_sfc()
lakes <- st_intersection(lakes, bbox_sf)

# union to compress
lakes <- st_union(lakes)

# plot
ggplot() +
  geom_sf(
    data = mudflats, fill = "#faf5ef", alpha = 0.6, colour = "#faf5ef"
  ) +
  geom_sf(data = land, fill = "#faf5ef", colour = "grey80") +
  geom_sf(
    data = lakes, fill = "#D7E7FF", colour = "grey80"
  ) +
  coord_sf(
    xlim = c(bbox["xmin"], bbox["xmax"]),
    ylim = c(bbox["ymin"], bbox["ymax"]),
    expand = FALSE
  ) +
  theme(
    panel.grid.major = element_line(colour = "transparent"),
    panel.grid.minor = element_line(colour = "transparent"),
    panel.background = element_rect(fill = "#D7E7FF"),
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.border = element_rect(fill = NA, colour = "grey20")
  )
```

### Save data

Includes data in the package, if `tools4watlas` is opened as project.

```{r, eval=FALSE}
# save data
save(mudflats, file = "../../data/muddflats.rda", compress = "xz")
save(lakes, file = "../../data/lakes.rda", compress = "xz")
```

