---
title: "Package maintenance"
author: "Johannes Krietsch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package maintenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  dpi = 300,
  message = FALSE
)
```

## Contributing to `tools4watlas`

If you want to contribute to `tools4watlas` [fork](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/fork-a-repo) the repository on GitHub and then submit a [pull request](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-pull-requests). Besides clear bug fixes, it is best to discuss potential changes or additions with [Allert Bijleveld](mailto:allert.bijleveld@nioz.nl). 

Before submitting a pull request to the original “upstream” repository, please make sure the code passes all R CMD checks (`devtools::check()`). If you changed content that changes the website, rebuild the website with `pkgdown::build_site()` into the docs folder. Note that some articles that are not vignettes, require a path to the WATLAS data folder "tools4watlas_articles" including additional data. 

For more details see [R Packages by Hadley Wickham and Jennifer Bryan](https://r-pkgs.org/) and read the documentation of the used packages. 

## Working on `tools4watlas`

When contributing, please follow the existing structure of the repository and [R style](http://adv-r.had.co.nz/Style.html). 

### Coding style

Follows the standard [tidyverse style](https://style.tidyverse.org/) and can be checked using [`lintr`](https://lintr.r-lib.org/). You can use [`styler`](https://styler.r-lib.org/) to format your code (will take care of most needed changes, but not all). Only use style in the file you are working on or selected code. The easiest is to use the R Studio addin otherwise use `styler::style_file("my_file.R")`.

Note that not all code follows this style at the moment, but new code should aim to be consistent with it. 

```{r, eval=FALSE}
# lintr tidyverse style
lintr::use_lintr(type = "tidyverse")

# in a project:
lintr::lint_dir()

# in a package:
lintr::lint_package()
```

### Documentation and functions

The package is documented useing [`roxygen2`](https://roxygen2.r-lib.org/). Functions are in the [R folder](https://github.com/allertbijleveld/tools4watlas/tree/master/R) and always have the prefix `atl_`. Each function has a [`roxygen2`](https://roxygen2.r-lib.org/) header. This header can be created by using `Ctrl + Shift + D` within a new function, if you’re using RStudio. Optimally, each function has an runable example. Once done run:

```{r, eval=FALSE}
devtools::document()
```

### Vignettes

To check if the vignettes are correctly compiled, either use the `Knit` button in R Studio or build all or single vignettes. This is also checked when running `devtools::check()`, so not necessary if everything is fine there (but can help when debugging).

```{r, eval=FALSE}
# check if all vignettes build
devtools::build_vignettes()

# check single vignette
devtools::build_rmd("vignettes/vignette-name.Rmd")
```

### Package website

The package website is build using [`pkgdown`](https://pkgdown.r-lib.org/). It's structure is defined in the [_pkgdown.yml](https://github.com/allertbijleveld/tools4watlas/blob/master/_pkgdown.yml). 

When editing anything simply run with `lazy = TRUE`. Otherwise, make sure you have to correct file path in articles that rely on local data. 

```{r, eval=FALSE}
# build site new without rebuilding all articles
pkgdown::build_site(lazy = TRUE)

# rebuilds everything (correct file path to local data?)
pkgdown::build_site()

```



### Check before pushing to GitHub

```{r, eval=FALSE}
# set working directory (not necessary if R Studio project)
# setwd("C:/Users/..path../tools4watlas")

# document
devtools::document()

# load all functions
devtools::load_all()

# build readme (run only if changed)
devtools::build_readme()

# tidy description
usethis::use_tidy_description()

# R CMD check
devtools::check()

# build website (lazy = TRUE only changes files that were edited)
pkgdown::build_site(lazy = TRUE)

```

Give meaningful commit message and commit the everything in the docs folder with the website separately (to make it easier to check changes).




